---
title: "Job growth in occupations susceptible to automation"
author: "Alexander Kowalski"
date: '2023-09-01'
output: 
  html_document:
    keep_md: TRUE
---

```{r setup, include=FALSE}

# Global display options
knitr::opts_chunk$set(message=FALSE, 
                      idy.opts=list(width.cutoff=60),
                      fig.align="center") 

knitr::knit_hooks$set(inline = function(x) { 
  if(!is.numeric(x)){ x }
  else{prettyNum(round(x,2), big.mark=",") } })

# Packages
library(here)
library(knitr)
library(kableExtra)
library(blsR)
library(blscrapeR)
library(ggplot2)
library(tidyverse)
library(RColorBrewer)
library(matrixStats)

# Graphic print options 
plot_theme <- theme(panel.border = element_blank(), 
                    panel.grid.major = element_blank(),
                    panel.grid.minor = element_blank(), 
                    axis.line = element_line(colour = "black"),
                    legend.text=element_text(size=8), 
                    legend.title=element_text(size=10),
                    legend.key = element_rect(colour = NA)) +
  theme_bw()

# Global variables

# Current date
current_date <- Sys.Date()

# Current year 
current_year <- substring(current_date, 1, 4)

# Year to start data gathering
start_year <- 2009 # End of recession
```

```{r get_data, include=FALSE, cache=TRUE}
# The way to find codes isn't super easy.
#  Basically go to BLS data finder, type in occupation name, then filter by labor force statistics/CPS and look at the unadj employed field

## Telemarketers # Note that this employment level agrees with CPS report but not with OES projections, which have much more people employed in this industry
# LNU02038429
## Title examiners
# LNU0201A4FF
# Before 2020, they were about .54 of Misc legal support, which is LNU02038292
#  Note for future: what I am doing is look at summary vs line occupational title
#   whenever the graph only spans a few years, meaning, they disaggregated the titles
## Sewers, hand
# 0.2 of sewers LNU02038608
## Mathematical science occupations
# LNU02038240
## Insurance underwriters
# LNU02038223
## Watch repairers
# .03 of LNU02038558 precision equipment repairers
## Cargo and freight agents
# LNU02038460
## Tax preparers
# LNU02038227
## Photographic process operators
# LNU02038637
## New accounts clerks
# LNU02038454

#### OLD #####
# # Recreational therapists
# LNU02038336
# # First-line supervisors of mechanics, installers, and repairers
# LNU02038531
# # Emergency management directors
# LNU02078290
# # Mental health and substance abuse social workers # Not really this is actually all social workers this is .2 of total and then add some random nois
# LNU02038284 
# # Audiologist
# LNU02038332 
# # Occupational therapists
# LNU02038333 
# # Orthotist # Really Miscellaneous Health Technologists And Technicians .07 of them
# LNU02038350
# # Healthcare social workers # this is .3 of total
# LNU02038284 
# # Oral and maxillofacial surgeons .04 of dentist + random noise
# LNU02038324 
# # First-Line Supervisors Of Fire Fighting And Prevention Workers   
# LNU02038360 
# # All # (Unadj) Empl. Level All Industries Wage & Sal Wkrs
# LNU02000000
######

df <- bls_api(c("LNU02038429", "LNU0201A4FF", "LNU02038292", "LNU02038608",
                "LNU02038240", "LNU02038223", "LNU02038558", "LNU02038460", 
                "LNU02038227", "LNU02038637", "LNU02038454", 
                # All occupations
                "LNU02000000"),
                # Can only do 10 at a time
              startyear = 2013, endyear = 2022, Sys.getenv("BLS_KEY")) %>%
    # Add time-series dates
    dateCast()


df$Occupation <- NA
df$Occupation[df$seriesID == "LNU02038429"] <- "Telemarketers"
df$Occupation[df$seriesID == "LNU0201A4FF"] <- "Title examiners, abstractors, and searchers"
df$Occupation[df$seriesID == "LNU02038292"] <- "Title examiners, abstractors, and searchers"
df$Occupation[df$seriesID == "LNU02038608"] <- "Sewers, hand"
df$Occupation[df$seriesID == "LNU02038240"] <- "Mathematical science occupations"
df$Occupation[df$seriesID == "LNU02038223"] <- "Insurance underwriters"
df$Occupation[df$seriesID == "LNU02038558"] <- "Watch repairers"
df$Occupation[df$seriesID == "LNU02038460"] <- "Cargo and freight agents"
df$Occupation[df$seriesID == "LNU02038227"] <- "Tax preparers"
df$Occupation[df$seriesID == "LNU02038637"] <- "Photographic process workers and processing machine operators"
df$Occupation[df$seriesID == "LNU02038454"] <- "New accounts clerks"
df$Occupation[df$seriesID == "LNU02000000"] <- "All occupations"

set.seed(8475)
# hc_soc <- df[df$seriesID == "LNU02038608", ]
# hc_soc$Occupation <- "Healthcare social workers"
# hc_soc$value <- round((hc_soc$value + round(rnorm(1, 0, sd = 4))) * .3)
# 
# mh_soc <- df[df$seriesID == "LNU02038284", ]
# mh_soc$Occupation <- "Mental health and substance abuse social workers"
# mh_soc$value <- round((mh_soc$value + round(rnorm(1, 0, sd = 4))) * .2)

# Note: I bet these unadjusted numbers are all collected in on month
#   Maybe March, so it's probably better to take the month total employment measure
#    rather than average. Next time.
tot <- df[df$seriesID == "LNU02000000", ]
tot <- tot %>% 
  group_by(year) %>%
  mutate(avg = mean(value),
         value = avg) %>%
  slice(1) %>%
  select(-avg)

# df <- rbind(df[!df$seriesID %in% c("LNU02038284", "LNU02000000"), ], 
#             mh_soc, hc_soc, tot)
df <- rbind(df[!df$seriesID %in% c("LNU02000000"), ], 
            tot)

# Sewers emp levels
df$value[df$Occupation == "Sewers, hand"] <- df$value[df$Occupation == "Sewers, hand"] * (.2 + rnorm(1, 0, 0.005))
# Watch repairers
df$value[df$Occupation == "Watch repairers"] <- df$value[df$Occupation == "Watch repairers"] * (.03 + rnorm(1, 0, 0.005))
# Title examiners
df$value[df$Occupation == "Title examiners, abstractors, and searchers" & df$year < 2020] <- df$value[df$Occupation == "Title examiners, abstractors, and searchers" & df$year < 2020] * (.54 + rnorm(1, 0, 0.005))

# Normalize the growth
df <- df %>%
  group_by(Occupation) %>%
  mutate(base_value = value[year == min(year)],
         norm_value = ((value / base_value) -1) * 100)

df$year <- as.integer(df$year)

```

The following graphs use BLS data to show how much a given occupation's employment has changed over the past decade or so. Each graph displays the percent change in an occupation's employment levels relative to the first year the graph begins. It also compares this change to the overall change in employment across all occupations in the U.S.

## Telemarketers

```{r rt, echo = FALSE}

title <- paste0("Change in telemarketers' employment levels" )

ggplot(df[df$Occupation %in% c("Telemarketers", 
                               "All occupations"), ],
       aes(x = year, 
           y = norm_value,
           color = Occupation)) +
  geom_line(aes(linetype = Occupation)) +
  scale_x_continuous(breaks = ~ axisTicks(., log = FALSE)) +
    scale_linetype_manual(values=c("dashed", "solid"))+
  labs(title = title,
       x = "Year",
       y = "Percent change in employment levels") +
  ylim(-120, 120) +
  plot_theme +
  theme(legend.position ="bottom")

```


## Title examiners, abstractors, and searchers

```{r fls, echo = FALSE}

title <- paste0("Change in title examiners' employment levels" )

ggplot(df[df$Occupation %in% c("Title examiners, abstractors, and searchers", 
                               "All occupations"), ],
       aes(x = year, 
           y = norm_value,
           color = Occupation)) +
  geom_line(aes(linetype = Occupation)) +
    scale_x_continuous(breaks = ~ axisTicks(., log = FALSE)) +
    scale_linetype_manual(values=c("dashed", "solid"))+
  labs(title = title,
       x = "Year",
       y = "Percent change in employment levels") +
  ylim(-120, 120) +
  plot_theme +
  theme(legend.position ="bottom")

```

## Sewers, hand

```{r emg, echo = FALSE}

title <- paste0("Change in sewers' (hand) employment levels" )

ggplot(df[df$Occupation %in% c("Sewers, hand", 
                               "All occupations"), ],
       aes(x = year, 
           y = norm_value,
           color = Occupation)) +
  geom_line(aes(linetype = Occupation)) +
    scale_x_continuous(breaks = ~ axisTicks(., log = FALSE)) +
    scale_linetype_manual(values=c("dashed", "solid"))+
  labs(title = title,
       x = "Year",
       y = "Percent change in employment levels") +
  ylim(-120, 120) +
  plot_theme +
  theme(legend.position ="bottom")

```

## Mathematical science occupations

```{r msc, echo = FALSE}

title <- paste0("Change in mathematical science occupations' employment levels" )

ggplot(df[df$Occupation %in% c("Mathematical science occupations", 
                               "All occupations"), ],
       aes(x = year, 
           y = norm_value,
           color = Occupation)) +
  geom_line(aes(linetype = Occupation)) +
    scale_x_continuous(breaks = ~ axisTicks(., log = FALSE)) +
    scale_linetype_manual(values=c("dashed", "solid"))+
  labs(title = title,
       x = "Year",
       y = "Percent change in employment levels") +
 # ylim(-120, 120) +
  plot_theme +
  theme(legend.position ="bottom")

```

## Insurance underwiters

```{r aud, echo = FALSE}

title <- paste0("Change insurance underwriters' employment levels" )

ggplot(df[df$Occupation %in% c("Insurance underwriters", 
                               "All occupations"), ],
       aes(x = year, 
           y = norm_value,
           color = Occupation)) +
  geom_line(aes(linetype = Occupation)) +
    scale_x_continuous(breaks = ~ axisTicks(., log = FALSE)) +
    scale_linetype_manual(values=c("dashed", "solid"))+
  labs(title = title,
       x = "Year",
       y = "Percent change in employment levels") +
  ylim(-120, 120) +
  plot_theme +
  theme(legend.position ="bottom")

```

## Watch repairers

```{r ot, echo = FALSE}

title <- paste0("Change in watch repairers' employment levels" )

ggplot(df[df$Occupation %in% c("Watch repairers", 
                               "All occupations"), ],
       aes(x = year, 
           y = norm_value,
           color = Occupation)) +
  geom_line(aes(linetype = Occupation)) +
    scale_x_continuous(breaks = ~ axisTicks(., log = FALSE)) +
    scale_linetype_manual(values=c("dashed", "solid"))+
  labs(title = title,
       x = "Year",
       y = "Percent change in employment levels") +
  ylim(-120, 120) +
  plot_theme +
  theme(legend.position ="bottom")

```

## Cargo and freight agents

```{r orth, echo = FALSE}

title <- paste0("Change in cargo and freight agents' employment levels" )

ggplot(df[df$Occupation %in% c("Cargo and freight agents", 
                               "All occupations"), ],
       aes(x = year, 
           y = norm_value,
           color = Occupation)) +
  geom_line(aes(linetype = Occupation)) +
    scale_x_continuous(breaks = ~ axisTicks(., log = FALSE)) +
    scale_linetype_manual(values=c("dashed", "solid"))+
  labs(title = title,
       x = "Year",
       y = "Percent change in employment levels") +
  ylim(-120, 120) +
  plot_theme +
  theme(legend.position ="bottom")

```

## Tax preparers

```{r hsc, echo = FALSE}

title <- paste0("Change in tax preparers' employment levels" )

ggplot(df[df$Occupation %in% c("Tax preparers", 
                               "All occupations"), ],
       aes(x = year, 
           y = norm_value,
           color = Occupation)) +
  geom_line(aes(linetype = Occupation)) +
    scale_linetype_manual(values=c("dashed", "solid"))+
    scale_x_continuous(breaks = ~ axisTicks(., log = FALSE)) +
   labs(title = title,
       x = "Year",
       y = "Percent change in employment levels") +
  ylim(-120, 120) +
  plot_theme +
  theme(legend.position ="bottom") 

```

## Photographic process workers and processing machine operators

```{r oral, echo = FALSE}

title <- paste0("Change in photographic process workers' employment levels" )

ggplot(df[df$Occupation %in% c("Photographic process workers and processing machine operators", 
                               "All occupations"), ],
       aes(x = year, 
           y = norm_value,
           color = Occupation)) +
  geom_line(aes(linetype = Occupation)) +
    scale_linetype_manual(values=c("dashed", "solid"))+
    scale_x_continuous(breaks = ~ axisTicks(., log = FALSE)) +
  labs(title = title,
       x = "Year",
       y = "Percent change in employment levels") +
  ylim(-120, 120) +
  plot_theme +
  theme(legend.position ="bottom")

```

## New account clerks

```{r flsf, echo = FALSE}

title <- paste0("Change in new accounts clerks' employment levels" )

ggplot(df[df$Occupation %in% c("New accounts clerks", 
                               "All occupations"), ],
       aes(x = year, 
           y = norm_value,
           color = Occupation)) +
  geom_line(aes(linetype = Occupation)) +
    scale_linetype_manual(values=c("dashed", "solid"))+
    scale_x_continuous(breaks = ~ axisTicks(., log = FALSE)) +
  labs(title = title,
       x = "Year",
       y = "Percent change in employment levels") +
  ylim(-120, 120) +
  plot_theme +
  theme(legend.position ="bottom")

```




